X-Factors: External Influences of a Hockey Goaltender's Performance

### 2024-12-04 paper/paper.qmd
saved prior reference edits

Few team sport athletes have as much influence on team performance as the hockey goaltender. Legendary performances, like Dominik Hasek single-handedly dragging Czechia past heavy favourites Canada and Russia to win the 1998 Olympic gold medal, are etched in the history books forever. Yet up until recently, it was hard to quantify goaltender performances; standard metrics such as wins, percentage of shots saved, and goals allowed per game did not tell the full story about goaltender's performance. The explosion in popularity of sports analytics has led to new metrics that better measure what fans see with the eye test. The most popular advanced metric, Goals Saved Above Expected (GSAx), is an all-encompassing statistic that captures a goaltender's performance relative to their peers. This is calculated by finding the difference between the goals a goaltender allows and the number a goaltender is 'expected' to allow; the expectation is quantified using an advanced model with many factors, such as MoneyPuck's model @moneypuck.

@@

To analyze the effect of external factors on goaltender performance, we use a log of 394,250 shots taken during the 2021-22 to 2024-25 NHL regular and post-seasons, provided by Moneypuck @moneypuck. Moneypuck is a publicly available source of hockey analytics that tracks advanced statistics, playoff odds, and event data. The shot data is updated daily from official NHL game reports, and is organized by season. For the current 2024-25 season, data is included up to November 26, 2024.

@@

Shots, defined as any player’s attempt to score by directing the puck at the net, are classified as an event type by the NHL. Events, along with their associated attributes and metadata, are officially recorded by Sportradar, the NHL's data partner, using AI-powered data tools @sportradar. Events are also counted and verified by official scorekeepers.

@@

- `shotslast3min:GSAx_so_far`  
   Coefficient: `0.002`  
   Implication: The interaction between recent activity (shots in the last three minutes) and cumulative performance (`GSAx_so_far`) suggests that recent activity amplifies the positive effects of prior performance.

@@

- Intercept: The baseline value itself is not particularly meaningful since most predictors will not realistically be zero.
- `isPlayoffGame`  
- `period2nd OT or later`  
- `period2nd Period`  
- `last_game_GSAx`  
- `last_5_avg_GSAx`  
- `isHomeTeam:isPlayoffGame`  
- `goalieTeamScoreDifferential:shotslast3min`

@@

We calculate variables required for our analysis model: Goals Saved Above Expected (`GSAx`) for each shot is obtained by subtracting the expected goal value of the shot `xGoal` from boolean value `goal`. `xGoal` is modelled by Moneypuck based on many other shot variables @moneypuck.

>>>>

Few team sport athletes have as much influence on team performance as the hockey goaltender. Legendary performances, like Dominik Hasek single-handedly dragging Czechia past heavy favourites Canada and Russia to win the 1998 Olympic gold medal, are etched in the history books forever. Yet up until recently, it was hard to quantify goaltender performances; standard metrics such as wins, percentage of shots saved, and goals allowed per game did not tell the full story about goaltender's performance. The explosion in popularity of sports analytics has led to new metrics that better measure what fans see with the eye test. The most popular advanced metric, Goals Saved Above Expected (GSAx), is an all-encompassing statistic that captures a goaltender's performance relative to their peers. This is calculated by finding the difference between the goals a goaltender allows and the number a goaltender is 'expected' to allow; the expectation is quantified using an advanced model with many factors, such as MoneyPuck's model [@moneypuck].

@@

To analyze the effect of external factors on goaltender performance, we use a log of 394,250 shots taken during the 2021-22 to 2024-25 NHL regular and post-seasons, provided by Moneypuck [@moneypuck]. Moneypuck is a publicly available source of hockey analytics that tracks advanced statistics, playoff odds, and event data. The shot data is updated daily from official NHL game reports, and is organized by season. For the current 2024-25 season, data is included up to November 26, 2024.

@@

Shots, defined as any player’s attempt to score by directing the puck at the net, are classified as an event type by the NHL. Events, along with their associated attributes and metadata, are officially recorded by Sportradar, the NHL's data partner, using AI-powered data tools [@sportradar]. Events are also counted and verified by official scorekeepers.

@@

- `shotslast3min * GSAx_so_far`  
   Coefficient: `0.002`  
   Implication: The interaction between recent activity (shots in the last three minutes) and cumulative performance (`GSAx_so_far`) suggests that recent activity amplifies the positive effects of prior performance.

@@

- Intercept: The baseline value itself is not particularly meaningful since most predictors will not realistically be zero.
- `isPlayoffGame`  
- `period2nd OT or later`  
- `period2nd Period`  
- `last_game_GSAx`  
- `last_5_avg_GSAx`  
- `isHomeTeam * isPlayoffGame` interaction
- `goalieTeamScoreDifferential * shotslast3min` interactoin

@@

We calculate variables required for our analysis model: Goals Saved Above Expected (`GSAx`) for each shot is obtained by subtracting the expected goal value of the shot `xGoal` from boolean value `goal`. `xGoal` is modelled by Moneypuck based on many other shot variables [@moneypuck].
